---
title: "Does the prescription of antibiotics commonly used to treat respiratory infections correlate with the number of COVID-19 cases?"
author: "Annie Brady"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)
```

```{r include=FALSE}
library(tidyverse)
library(janitor) # cleaning data
library(here) # for help importing data
library(knitr) #for making tables
library(kableExtra) #for improving my tables
library(sf) #for making the map plot
```
## Introduction
The data used in this report can be found at Public Health Scotland:
https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community

In this report, I will explore the relationship between the prescription of nine antibiotics commonly prescribed on the NHS to treat respiratory infections, and the number of COVID-19 cases in Scotland in 2021.

Antimicrobial resistance (AMR) is a growing global concern, and one of the key contributors to its rise is the overuse and misuse of antibiotics. In this context, it is crucial to understand how patterns of antibiotic prescribing may be influenced by factors such as the COVID-19 pandemic. Specifically, I aim to investigate whether the prescription rates of these antibiotics rose and fell with the number of COVID-19 cases, potentially indicating the prescription of antibiotics to people suffering from COVID, not bacterial infections. It is important to note that COVID-19 is caused by a virus, and antibiotics are not effective in treating viral infections. Therefore, any rise in prescriptions could reflect inappropriate use of antibiotics, which may contribute to the AMR problem.

### Prescription Data
```{r}
data_jan2021 <- read.csv(here("data", "data_jan2021.csv")) %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, c("PHENOXYMETHYLPENICILLIN|AMOXICILLIN| CO_AMOXICLAV|AZITHROMYCIN|CLARITHROMYCIN|ERYTHROMYCIN|VIBRAMYCIN|DOXYCYCLINE|CEFALEXIN"))) %>% 
#Using the 'clean_names' function to clean up the column names by adding underscores and making them all lowercase. I then use the 'str_detect' command to filter for any rows containing the above drug names in the column "bnf_item_description".
    filter(paid_quantity == floor(paid_quantity))
```
I'm going to create a function to simplify the import, cleaning and filtering of my data, as I will have to do the same thing to each month. I have added a line of code to remove values that are not whole number from the 'paid_quantity' column. This is because some of the datasets have errors where this column has a decimals, which is not possible as it is not possible to give 0.5 of a prescription.

Talk about the Drugs I'm filtering for and why: Phenoxymethylpenicillin, Amoxicillin, co-Amoxiclav, Azithromycin, Clarithromycin, Erythromycin, Vibramycin-D (doxycycline), Doxycycline, Cefalexin.
```{r}
function_clean_filter_drugs <- function(data_month) {
    read_csv(here("data", data_month), show_col_types = FALSE) %>% 
    clean_names() %>% 
    filter(str_detect(bnf_item_description, c("PHENOXYMETHYLPENICILLIN|AMOXICILLIN| CO_AMOXICLAV|AZITHROMYCIN|CLARITHROMYCIN|ERYTHROMYCIN|VIBRAMYCIN|DOXYCYCLINE|CEFALEXIN"))) %>% 
    filter(paid_quantity == floor(paid_quantity)) #floor rounds down to the nearest whole number, so this line filters for values in 'paid_quantity' that have no decimals.
}
```
Now I am going to download the datasets for each month in 2021, so that I can plot a graph of the correlation between antibiotic prescriptions and covid cases in 2021.
```{r}
data_feb2021 <- function_clean_filter_drugs("data_feb2021.csv")
data_march2021 <- function_clean_filter_drugs("data_march2021.csv")
data_april2021 <- function_clean_filter_drugs("data_april2021.csv")
data_may2021 <- function_clean_filter_drugs("data_may2021.csv")
data_june2021 <- function_clean_filter_drugs("data_june2021.csv")
data_july2021 <- function_clean_filter_drugs("data_july2021.csv")
data_aug2021 <- function_clean_filter_drugs("data_aug2021.csv")
data_sep2021 <- function_clean_filter_drugs("data_sep2021.csv")
data_oct2021 <- function_clean_filter_drugs("data_oct2021.csv")
data_nov2021 <- function_clean_filter_drugs("data_nov2021.csv")
data_dec2021 <- function_clean_filter_drugs("data_dec2021.csv")
```
Now I will wrangle these to give me totals for each healthboard. I have chosen to do total in each health board, firslt so i can easily join the datasets, but also so that later on I can compare the rates of antibiotic presciption in each healthboard, and see if there are any that are especially bad for overprescripton.
```{r}
jan_2021_totals <- data_jan2021 %>% 
  group_by(hbt) %>% 
    summarise(january = sum(paid_quantity)) %>% 
  filter(hbt != "SB0806")
feb_2021_totals <- data_feb2021 %>% 
  group_by(hbt) %>% 
    summarise(february = sum(paid_quantity))
march_2021_totals <- data_march2021 %>% 
  group_by(hbt) %>% 
    summarise(march = sum(paid_quantity))
april_2021_totals <- data_april2021 %>% 
  group_by(hbt) %>% 
    summarise(april = sum(paid_quantity))
may_2021_totals <- data_may2021 %>% 
  group_by(hbt) %>% 
    summarise(may = sum(paid_quantity))
june_2021_totals <- data_june2021 %>% 
  group_by(hbt) %>% 
    summarise(june = sum(paid_quantity))
july_2021_totals <- data_july2021 %>% 
  group_by(hbt) %>% 
    summarise(july = sum(paid_quantity))
aug_2021_totals <- data_aug2021 %>% 
  group_by(hbt) %>% 
    summarise(august = sum(paid_quantity))%>%  
  filter(hbt != "SB0806")#removing an error - this health board does not exist
sep_2021_totals <- data_sep2021 %>% 
  group_by(hbt) %>% 
    summarise(september = sum(paid_quantity))
oct_2021_totals <- data_oct2021 %>% 
  group_by(hbt) %>% 
    summarise(october = sum(paid_quantity)) %>%  
  filter(hbt != "SB0806") 
nov_2021_totals <- data_nov2021 %>% 
  group_by(hbt) %>% 
    summarise(november = sum(paid_quantity))%>%  
  filter(hbt != "SB0806")
dec_2021_totals <- data_dec2021 %>% 
  group_by(hbt) %>% 
    summarise(december = sum(paid_quantity))
```
Now join them all together:
```{r}
prescriptions_2021 <- jan_2021_totals %>% 
  full_join(feb_2021_totals, by = "hbt") %>% 
  full_join(march_2021_totals, by = "hbt") %>% 
  full_join(april_2021_totals, by = "hbt") %>% 
  full_join(may_2021_totals, by = "hbt") %>% 
  full_join(june_2021_totals, by = "hbt") %>% 
  full_join(july_2021_totals, by = "hbt") %>% 
  full_join(aug_2021_totals, by = "hbt") %>% 
  full_join(sep_2021_totals, by = "hbt") %>% 
  full_join(oct_2021_totals, by = "hbt") %>% 
  full_join(nov_2021_totals, by = "hbt") %>% 
  full_join(dec_2021_totals, by = "hbt")
  
prescriptions_monthly_totals <- prescriptions_2021 %>% 
  select(-hbt) %>% 
  pivot_longer(cols = c("january":"december"),
               names_to = "month",
               values_to = "prescriptions") %>% 
  group_by(month) %>% 
  summarise(prescriptions = sum(prescriptions))
```
### COVID-19 Data
Now I want to load the covid data from Public Health Scotland - this data set is "Weekly Tests and Cases by Health Board" from the "Viral Respiratory Diseases (Including Influenza and COVID-19) Data in Scotland" colleciton. I chose this set so that I can join it to my other data by healthboard.
```{r}
covid_cases_tests_hb <- read_csv(here("data", "covid_cases_tests_hb.csv"), show_col_types = FALSE) %>% 
  clean_names() %>% 
  filter(hb != "S92000003") %>% #removed the aggregated data for the whole of Scotland
  mutate(week_ending = ymd(week_ending)) #convert the week_ending column from dbl to date for easier analysis.
```
Filter for the year 2021 and then turn weekly cases into monthly cases - create a graph tracking cases?
```{r}
covid_monthly_total_2021 <- covid_cases_tests_hb %>% 
  filter(year(week_ending) == 2021) %>% #filter for only the year 2021
  mutate(month = month(week_ending)) %>% #create a column containing the month so that group_by can be applied 
  group_by(month) %>% 
  summarise(monthly_cases = sum(weekly_cases)) %>% #add up the weekly cases, presenting total per month.
    mutate(month = recode(month, "1" = "january","2" = "february","3" = "march","4" = "april","5" = "may","6" = "june","7" = "july","8" = "august", "9" = "september","10" = "october","11" = "november","12" = "december"))
```
### Join COVID and Prescription Data
Here I joined `prescriptions_2021` to `covid_monthly_total_2021`, so that I can compare COVID cases and prescription rates.
```{r}
prescriptions_covid_2021 <- prescriptions_monthly_totals %>%
  full_join(covid_monthly_total_2021, by = "month") %>% #the months are lowercase and out of order, so this needs fixing.
  mutate(month = str_to_title(month)) %>%  #coverts months to title case, so that month.name can be applied.
  mutate(month = factor(month, levels = month.name, ordered = TRUE)) %>% #this coverts the month column to factors with sepcified levels. month.name is a built in character vector that contains the names of the months in order.
  arrange(month) #arrange the months in order
```
Here I use `geom_point` and `geom_smooth` from `ggplot()` to generate a graph comparing Antibiotic prescriptions and COVID cases.
```{r}
prescriptions_covid_2021 %>% 
  ggplot(aes(x = prescriptions,
             y = monthly_cases)) +
  geom_point()+
  geom_smooth(method = "lm", color = "lightpink", linetype = "solid", size = 1) + #method = lm instructs geom_smooth to use linear regression to make a line that represents the relationship between x and y.
  labs(title = "Antibiotic Prescriptions* vs. COVID Cases in 2021", x = "Number of Prescriptions", y = "COVID Cases",
       caption = "* Antibiotics commonly prescribed for resipiritory infections") +
  theme_minimal()
```
This graph shows a positive correlation between the number of antibiotic prescriptions and COVID cases. he grey area represents the 95% confidence interval for the fitted line.I wanted to get more evidence for this correlation, so I used the Pearson Correlation test.
```{r}
cor.test(prescriptions_covid_2021$prescriptions, prescriptions_covid_2021$monthly_cases, method = "pearson")
```
The p-value is < 0.05, meaning I can reject the null hypothesis, suggesting a statistically significant correlation. cor = 0.613, suggesting positive correlation. This corroborates the correlation that can be seen on the graph.  
Plan for the rest of the assignment:    
   
Aggregate the prescriptions data by health board and see if there is health board that was especially bad for it (divide total antibiotic prescriptions by the number of people living in that healthboard) and compare. Visualise on a map.

Possibly compare prescription rate to a non-COVID year (2019) if space allows.
